Convex Surrogates via Polyhedral Losses

### 2019-02-01 cvx-fin.tex
pass through section 4

Generalizing the above to higher dimensions, we obtain the following.
\begin{theorem}
  A finite property $\gamma:\simplex\toto\R$ is $d$-embeddable if and only if there exist polytopes $T(r,y) \subseteq \reals^d$ for all $r\in\R, y\in\Y$, such that the following two conditions hold:
  \begin{enumerate}
  \item (Optimality) For all $r\in\R, p\in\simplex$, we have $r\in \gamma(p) \iff 0 \in \sum_{y\in\Y} p_y T(r,y)$, where summation is the Minkowski sum.
  \item (Monotonicity) There exists an embedding $\varphi: \R \to \reals^d$ and set of convex functions $\{L_y:\reals^d\to\reals\}_{y\in\Y}$ such that for all $y\in\Y, r\in\R$ we have $T(r,y) = \partial L_y(\varphi(r))$.
  \end{enumerate}
\end{theorem}

\bo{Added the below blurb}
\raf{I love it.  This would be a great place to discuss the relationship to FSD, which is ``even more local'' in the sense that it does not even look at a whole cell at once, just one piece of the boundary.  We can then say that the additional power is strong enough to give novel lower bounds where FSD remains silent.  (In particular, abstain looks 1d / orderable when zoomed in as far as FSD, but one can conclude it is not orderable using condition 1 alone.)}
This characterization decomposes embeddability into \emph{local} and \emph{global} necessary and sufficient conditions.
The local conditions at each embedding point turn out to imply very strong restrictions on the structure of possible polyhedral losses, regardless of any other structure in the property or where the embedding points are placed.
Meanwhile, the global condition says that, once each local embedding point's demands are satisfied, we must be able to place them all in $\reals^d$ in a coherent manner that simultaneously respects all of their requirements.

\bo{I think we should write separately the result that the expected loss function only has vertices above the embedding points (um, sort of - WLOG?). I think we've known it for so long we forgot that it's a really interesting and also a necessary and sufficient condition. Does it also imply that each individual loss function's vertices can only be at embedding points?}
\raf{This is one of those statements that is probably true in some sense but we're not sure what sense.  There could be vertices that are just never optimal, and it's hard to know exactly how to rule them out.}

\subsection{Useful necessary conditions}

@@

\begin{corollary}\label{cor:d-embed-implies-d-construction}
  If a property is $d$-embeddable, then each of its level sets has a $d$-construction.
  \bo{Is it possible that some uniqueness of the construction is also true?}\jessie{Up to rotation, scaling, and translation, possible but I wouldn't think so.  ex: Abstain(1/2) on 3 outsomes can be embedded in $\reals^2$ with both an equilateral and isosceles triangle}
\end{corollary}
\begin{proof}
  Suppose the level set is represented as $\Gamma_r = \{p : Bp \geq 0\}$ with $B \in \reals^{k \times n}$ for the minimum $k$ (i.e. the number of facets of the polytope).
  Let $u \in \reals^d$ be its embedding point.
  The convex expected loss is minimized at $u$ if and only if it has the zero vector as a subgradient at $u$, i.e.
    \[ p \in \Gamma_r \qquad \iff \qquad 0 \in \partial \langle L(u), p \rangle . \]
  Subgradient sets satisfy the following two properties: $\partial(f_1 + f_2) = \partial f_1 + \partial f_2$ where the right side is the Minkowski sum of the subgradient sets (i.e. the set of sums of all pairs); and $\partial(\alpha f_1) = \alpha \partial f_1$.
  So
    \[ \partial \langle p , L(u) \rangle = \sum_{y \in \Y} p_y \partial L(u)_y \]
  where the right side is a Minkowski sum of subgradient sets.
  We now consider the halfspaces defining this set.
  Define the $y$th support function as $H_y(v) = \max_{w \in \partial L(u)_y} \langle w, v \rangle$ to obtain that $w \in \partial \langle p, L(u) \rangle$ if and only if, for all unit vectors $v$,
   \[ \langle v, w \rangle \leq \sum_{y \in \Y} p_y H_y(v) . \]
  In particular, the embedding point is optimal for $p$ if and only if this is satisfied for $w=0$.
  So, letting $B_i$ be the $i$th column of $B$, we have
    \[ \Gamma_r = \{p : \sum_y p_y B_y \geq 0\} = \{p : \sum_y p_y H_y(v) \geq 0 ~ (\forall v, \|v\|=1)\} . \]
  These sets are equal and full-dimensional.
  Therefore, \bo{cite please!}, there exist $k$ vectors $v_1,\dots,v_k \in \reals^d$ such that $H_y(v_i) = B_{iy}$.
  Collected as rows of $V$, we obtain that each $(V,B_y)$ is a tight halfspace representation, as it contains $\partial L(u)_y$ and $H_y(v_i) = B_{iy}$, meaning there exists $w \in \partial L(u)_y$ such that $\langle w, v_i \rangle = B_{iy}$.
\end{proof}
This provides the following recipe for both constructing losses and proving lower bounds.
First, take any level set and represent it as $\{p : Bp \geq 0\}$.
Now if $B$ has $k$ rows, search for a collection of $k$ vectors in $\reals^d$.
For any candidate $V$ (whose rows are the vectors), a computer program can automatically verify whether the representations $\{(V,B_y)\}_{y \in \Y}$ are tight.
In particular, if any $(V,B_y)$ is infeasible (i.e. there does not exist any $w$ satisfying all constraints $\langle v_i, w \rangle \leq B_{iy}$), then $V$ is not a $d$-construction.
If $(V,B_y)$ is feasible but some constraints are redundant, i.e. it is not tight, then again we do not have a $d$-construction.

@@

\raf{Paragraph explaining two other useful necessary conditions: WMON, and reduction to 1 dimension}

\subsection{Examples}

\begin{theorem}[\bo{cite}]
  \bo{Still not proven}
  The mode with $|\Y|=n$ is not $(n-2)$-embeddable.
\end{theorem}
\begin{proof}
  Call the outcomes $\Y = \{1,\dots,n\} = \R$.
  We will pick the level set for the report $r=n$ and show it has no $d$-construction.
  There are two kinds of constraints on the level set.
  The constraints that $n$ be the mode implies $p_n \geq p_i ~ (\forall i \neq n)$.
  These can be rearranged as $-p_i + p_n \geq 0$.
  Then there are nonnegativity constraints on each $p_i$.
  This is redundant for $p_n$ so cannot be included.
  \[ B = \begin{pmatrix}
        -1 &  0 & \cdots & 0  & 1 \\
         0 & -1 & \cdots & 0  & 1 \\
           &    & \vdots &    &   \\
         0 &  0 & \cdots & -1 & 1  \\
         \hline
         1 &  0 & \cdots &    & 0  \\
         0 &  1 & \cdots &    & 0  \\
           &    & \vdots &    &    \\
         0 &    &        & 1  & 0
     \end{pmatrix} .
  \]
  Here $B$ has $2(n-1)$ rows, which we will call $1,2,\dots,n-1,1',2',\dots,(n-1)'$.

  A $d$-construction is therefore a $V$ with $n-1$ rows, each a vector $v_i$ in $\reals^d$.
  Suppose $d \leq n-2$, i.e. there are at least $d+1$ rows.
  Now consider the halfspace representation given by the second column of $B$, namely $T = \{ w : Vw \leq (-1, 0, \dots, 0)^{\intercal}\}$.
  The first $d+1$ rows are linearly dependent in $\reals^d$.
  Therefore, we can write the $d+1$st row as $v_{d+1} = \alpha_1 v_1 + \cdots + \alpha_d v_d$.
  So if $w \in T$, then $\inprod{v_1}{w} \leq -1$ and $\inprod{v_i}{w} \leq 0$ for all $i=2,\dots,d$.
  Taking the weighted combination, we obtain $\inprod{v_{d+1}}{w} \leq -\alpha_1$.
  If $\alpha_1$ is positive, this implies there is slack in the $d+1$st constraint, so the halfspace representation cannot be tight.
  If $\alpha_1$ is nonpositive, then we pick some other pair of vectors $i \neq j$ rather than $1,d+1$, such that $v_j = \sum_{i'\neq j} \alpha_{i'} v_{i'}$ with $\alpha_i > 0$.
  \bo{I claim there must exist two such vectors in this collection ... need to prove it though!}
  Choosing the $i$th column of $B$ as a constraint gives the contradiction as above.
\end{proof}
To recall what is happening inside this proof: we chose the first level set, which gave us a matrix $B$.
The first column was all-$1$, which corresponds to a constraint on the subgradient set of the loss on outcome $1$ for report $1$, which is not an interesting constraint.
The other constraints gave a contradiction if $d$ is too small.
\bo{More to say?}

\begin{theorem}\label{thm:abs-5outcome-not-2embed}
  Abstain with $\alpha=1/2$ and $|\Y|=5$ is not $2$-embeddable.
\end{theorem}
\begin{proof}
By Corollary~\ref{cor:d-embed-implies-d-construction}, we can show there is no $2$-construction for one of the level sets.
We choose the abstain cell, which is $\{p : \max_y p_y \leq 0.5\}$.
Because this equals $\{p : (\forall y) 2p_y \geq \sum_{y'} p_{y'}$\}, we can write the level set as $\{p : Bp \geq 0\}$ for
  \[ B = \begin{pmatrix}
       -1 &  1 &  1 & \cdots & 1  \\
       1  & -1 &  1 & \cdots & 1  \\
          &    &    & \vdots &    \\
       1  &  1 &  1 & \cdots & -1
     \end{pmatrix} .
  \]
In other words, $B := \ones - 2I$, where $\ones$ is the all-ones matrix and $I$ is the identity.
In our example, $|\Y| = 5$, so $B$ has width $5$ because there are $5$ outcomes and height $5$ because there are $5$ constraints (one per outcome).

@@

\bo{I think a figure would be really great here. One showing the 5 vectors, and one showing the constraints we get from one column of $B$ (i.e. the triangle.}

@@

\paragraph{On the matching upper bound.}
In constrast to this lower bound, \cite{ramaswamy2018consistent} give a $2$-embedding of the abstain loss on $|\Y|=4$.
This implies that a $2$-construction for the $4 \times 4$ version of the above $B$ matrix exists.
Indeed, one can take $V$ to be the four unit vectors pointing directly north, south, east, and west; the H-representations are then line segments, e.g. the segment $w_1=-1, w_2 \in [-1,1]$, etc.
These are all tight.
Indeed, as required by our characterization, the subgradient sets of their loss functions at the abstain embedding point span these H-reps (in fact are equal to them).

\bo{I think we should actually construct the losses for the ``rhombus'' and show that it embeds abstain loss. That might be a cool showcase for our understanding.}

>>>>

Generalizing the above to higher dimensions, we obtain the following.
\begin{theorem} \label{thm:char}
  A finite property $\gamma:\simplex\toto\R$ is $d$-embeddable if and only if there exist polytopes $T(r,y) \subseteq \reals^d$ for all $r\in\R, y\in\Y$, such that the following two conditions hold:
  \begin{enumerate}
  \item (Optimality) For all $r\in\R, p\in\simplex$, we have $r\in \gamma(p) \iff 0 \in \sum_{y\in\Y} p_y T(r,y)$, where summation is the Minkowski sum.
  \item (Monotonicity) There exists an embedding $\varphi: \R \to \reals^d$ and set of convex functions $\{L_y:\reals^d\to\reals\}_{y\in\Y}$ such that for all $y\in\Y, r\in\R$ we have $T(r,y) = \partial L_y(\varphi(r))$.
  \end{enumerate}
\end{theorem}

%\bo{Added the below blurb}
%\raf{I love it.  This would be a great place to discuss the relationship to FSD, which is ``even more local'' in the sense that it does not even look at a whole cell at once, just one piece of the boundary.  We can then say that the additional power is strong enough to give novel lower bounds where FSD remains silent.  (In particular, abstain looks 1d / orderable when zoomed in as far as FSD, but one can conclude it is not orderable using condition 1 alone.)}
This characterization decomposes embeddability into \emph{local} and \emph{global} necessary and sufficient conditions.
The local conditions at each embedding point turn out to imply very strong restrictions on the structure of possible polyhedral losses, regardless of any other structure in the property or where the embedding points are placed.
Meanwhile, the global condition says that, once each local embedding point's demands are satisfied, we must be able to place them all in $\reals^d$ in a coherent manner that simultaneously respects all of their requirements.

For one-dimensional embeddings, the monotonicity property is extremely useful for lower bounds, as it is easy to check whether or not a given property is orderable.
However, for higher dimensions, we suggest that the optimality condition is also a promising direction for lower bounds.
To that end, we next derive a more specific necessary condition and use it to prove a lower bound on the ``abstain loss''.

%\bo{I think we should write separately the result that the expected loss function only has vertices above the embedding points (um, sort of - WLOG?). I think we've known it for so long we forgot that it's a really interesting and also a necessary and sufficient condition. Does it also imply that each individual loss function's vertices can only be at embedding points?}
%\raf{This is one of those statements that is probably true in some sense but we're not sure what sense.  There could be vertices that are just never optimal, and it's hard to know exactly how to rule them out.}

\subsection{Necessary condition: $d$-construction}

@@

\begin{corollary}\label{cor:d-embed-implies-d-construction}
  If a property is $d$-embeddable, then each of its level sets has a $d$-construction.
\end{corollary}
\begin{proof}
  Suppose the level set is represented as $\Gamma_r = \{p : Bp \geq 0\}$ with $B \in \reals^{k \times n}$ for the minimum $k$ (i.e. the number of facets of the polytope).
  Let $u \in \reals^d$ be its embedding point.
  The convex expected loss is minimized at $u$ if and only if it has the zero vector as a subgradient at $u$, i.e.
    \[ p \in \Gamma_r \qquad \iff \qquad 0 \in \partial \langle L(u), p \rangle . \]
  Subgradient sets satisfy the following two properties: $\partial(f_1 + f_2) = \partial f_1 + \partial f_2$ where the right side is the Minkowski sum of the subgradient sets (i.e. the set of sums of all pairs); and $\partial(\alpha f_1) = \alpha \partial f_1$.
  So
    \[ \partial \langle p , L(u) \rangle = \sum_{y \in \Y} p_y \partial L(u)_y \]
  where the right side is a Minkowski sum of subgradient sets.
  We now consider the halfspaces defining this set.
  Define the $y$th support function as $H_y(v) = \max_{w \in \partial L(u)_y} \langle w, v \rangle$ to obtain that $w \in \partial \langle p, L(u) \rangle$ if and only if, for all unit vectors $v$,
   \[ \langle v, w \rangle \leq \sum_{y \in \Y} p_y H_y(v) . \]
  In particular, the embedding point is optimal for $p$ if and only if this is satisfied for $w=0$.
  So, letting $B_i$ be the $i$th column of $B$, we have
    \[ \Gamma_r = \{p : \sum_y p_y B_y \geq 0\} = \{p : \sum_y p_y H_y(v) \geq 0 ~ (\forall v, \|v\|=1)\} . \]
  These sets are equal and full-dimensional.
  Therefore, \bo{cite please!}, there exist $k$ vectors $v_1,\dots,v_k \in \reals^d$ such that $H_y(v_i) = B_{iy}$.
  Collected as rows of $V$, we obtain that each $(V,B_y)$ is a tight halfspace representation, as it contains $\partial L(u)_y$ and $H_y(v_i) = B_{iy}$, meaning there exists $w \in \partial L(u)_y$ such that $\langle w, v_i \rangle = B_{iy}$.
\end{proof}
This provides the following recipe for both constructing losses and proving lower bounds.
First, take any level set and represent it as $\{p : Bp \geq 0\}$.
Now if $B$ has $k$ rows, search for a collection of $k$ vectors in $\reals^d$.
For any candidate $V$ (whose rows are the vectors), a computer program can automatically verify whether the representations $\{(V,B_y)\}_{y \in \Y}$ are tight.
In particular, if any $(V,B_y)$ is infeasible (i.e. there does not exist any $w$ satisfying all constraints $\langle v_i, w \rangle \leq B_{iy}$), then $V$ is not a $d$-construction.
If $(V,B_y)$ is feasible but some constraints are redundant, i.e. it is not tight, then again we do not have a $d$-construction.

@@

\paragraph{Other necessary conditions.}
One can also derive other kinds of potentially necessary conditions from the general characterization in Theorem \ref{thm:char}.
One is related to the weak monotonicity property of convex functions.\bo{raf please fill in}

Another is a reduction to the one-dimensional characterization. If one considers any embedded property in $\reals^d$ and restricts the report space to a line while keeping the loss the same, one obtains a new $1$-dimensional embedded property.
Picking a line that e.g. goes through previous embedding or indifference points, those that were optimal originally must still be optimal under the same distribution.
At this point, the one-dimensional characterization may rule out embeddability.

\subsection{Example}
We give an example of how this necessary condition can be applied to prove a lower bound.
The \emph{abstain property} sets a threshold of $\alpha$ (here, $\alpha=0.5$) and has $\R = \Y \cup \{\text{abstain}\}$.
We have $\text{abstain} \in \gamma(p) \iff \max_y p_y \leq \alpha$, and $y \in \gamma(p) \iff p_y = \max_{y'} p_{y'}, p_y \geq \alpha$.
In other words, it is similar to the mode, except that if none of the options has a high enough probability, the best report is to abstain.
\citet{ramaswamy2018consistent} show that this property on $|\Y|=n$ can be embedded in $\lceil \log_2(n) \rceil$ dimensions.
Their construction is quite beautiful and extremely instructive for studying embeddings.
Prior to this work, it was not known if one could improve by embedding abstain($0.5$) on $5$ or more outcomes in $\reals^2$.

%\begin{theorem}[\bo{cite}]
%  \bo{Still not proven}
%  The mode with $|\Y|=n$ is not $(n-2)$-embeddable.
%\end{theorem}
%\begin{proof}
%  Call the outcomes $\Y = \{1,\dots,n\} = \R$.
%  We will pick the level set for the report $r=n$ and show it has no $d$-construction.
%  There are two kinds of constraints on the level set.
%  The constraints that $n$ be the mode implies $p_n \geq p_i ~ (\forall i \neq n)$.
%  These can be rearranged as $-p_i + p_n \geq 0$.
%  Then there are nonnegativity constraints on each $p_i$.
%  This is redundant for $p_n$ so cannot be included.
%  \[ B = \begin{pmatrix}
%        -1 &  0 & \cdots & 0  & 1 \\
%         0 & -1 & \cdots & 0  & 1 \\
%           &    & \vdots &    &   \\
%         0 &  0 & \cdots & -1 & 1  \\
%         \hline
%         1 &  0 & \cdots &    & 0  \\
%         0 &  1 & \cdots &    & 0  \\
%           &    & \vdots &    &    \\
%         0 &    &        & 1  & 0
%     \end{pmatrix} .
%  \]
%  Here $B$ has $2(n-1)$ rows, which we will call $1,2,\dots,n-1,1',2',\dots,(n-1)'$.
%
%  A $d$-construction is therefore a $V$ with $n-1$ rows, each a vector $v_i$ in $\reals^d$.
%  Suppose $d \leq n-2$, i.e. there are at least $d+1$ rows.
%  Now consider the halfspace representation given by the second column of $B$, namely $T = \{ w : Vw \leq (-1, 0, \dots, 0)^{\intercal}\}$.
%  The first $d+1$ rows are linearly dependent in $\reals^d$.
%  Therefore, we can write the $d+1$st row as $v_{d+1} = \alpha_1 v_1 + \cdots + \alpha_d v_d$.
%  So if $w \in T$, then $\inprod{v_1}{w} \leq -1$ and $\inprod{v_i}{w} \leq 0$ for all $i=2,\dots,d$.
%  Taking the weighted combination, we obtain $\inprod{v_{d+1}}{w} \leq -\alpha_1$.
%  If $\alpha_1$ is positive, this implies there is slack in the $d+1$st constraint, so the halfspace representation cannot be tight.
%  If $\alpha_1$ is nonpositive, then we pick some other pair of vectors $i \neq j$ rather than $1,d+1$, such that $v_j = \sum_{i'\neq j} \alpha_{i'} v_{i'}$ with $\alpha_i > 0$.
%  \bo{I claim there must exist two such vectors in this collection ... need to prove it though!}
%  Choosing the $i$th column of $B$ as a constraint gives the contradiction as above.
%\end{proof}

\bo{A whole section of the following proof can be cut out, but it's trying to be a tutorial. Pairs well with the figure (do we have a figure?)}
\begin{theorem}\label{thm:abs-5outcome-not-2embed}
  Abstain with $\alpha=1/2$ and $|\Y|=5$ is not $2$-embeddable.
\end{theorem}
\begin{proof}
By Corollary~\ref{cor:d-embed-implies-d-construction}, we can show there is no $2$-construction for one of the level sets.
We choose the abstain cell, which is $\{p : \max_y p_y \leq 0.5\}$.
Because this equals $\{p : (\forall y) 2p_y \geq \sum_{y'} p_{y'}$\}, we can write the level set as $\{p : Bp \geq 0\}$ for
  \[ B = \begin{pmatrix}
       -1 &  1 &  1 & \cdots & 1  \\
       1  & -1 &  1 & \cdots & 1  \\
          &    &    & \vdots &    \\
       1  &  1 &  1 & \cdots & -1
     \end{pmatrix} .
  \]
In other words, $B := \ones - 2I$, where $\ones$ is the all-ones matrix and $I$ is the identity.
In our example, $|\Y| = 5$, so $B$ has width $5$ because there are $5$ outcomes and height $5$ because there are $5$ constraints (one per outcome).

@@

@@

\paragraph{On the matching upper bound.}
One cannot prove a better lower bound for abstain loss and $\reals^2$, as \citet{ramaswamy2018consistent} give a $2$-embedding for $|\Y|=4$.
This implies that a $2$-construction for the $4 \times 4$ version of the above $B$ matrix exists; and indeed, the above proof roughly reveals its shape.
One can take $V$ to be the four unit vectors pointing directly north, south, east, and west; the H-representations are then line segments, e.g. the segment $w_1=-1, w_2 \in [-1,1]$, etc.
These are all tight.
Indeed, as required by our characterization, the subgradient sets of their loss functions at the abstain embedding point span these H-reps (in fact are equal to them).

\bo{It would be nice to actually construct the losses for the ``rhombus'' and show that it embeds abstain loss. That might be a cool showcase for our understanding.}
